---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)
library(modelsummary)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
datasummary_skim(gapminder)
summary(gapminder)
gapminder
```

**Observations**:

- Country, continent, year, lifeExp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(gapminder["year"])
year_min <- min(gapminder["year"])
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder_min_year <- gapminder |> 
  filter(year == year_min)

gapminder_min_means <- aggregate(
  gdpPercap ~ continent, 
  data = gapminder_min_year, 
  FUN = mean
  )

gapminder_min_means |> 
  ggplot() +
  geom_col(
    fill = "sky blue",
    aes(
      x = continent,
      y = gdpPercap
    )
  ) +
  labs(
    x = "Continent", 
    y = "Average GDP Per Capita ($USD)",
    title = "Average GDP Per Capita vs. Continent in 1952"
    ) +
  theme(plot.title = element_text(hjust = 0.5)) 


gapminder_min_year |> 
  ggplot() +
  geom_point(
    aes(
      x = continent,
      y = gdpPercap
    )
  ) +
  labs(
    x = "Continent", 
    y = "Average GDP Per Capita ($USD)",
    title = "Average GDP Per Capita vs. Continent in 1952"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) 


```

**Observations**:

- Oceania has a drastically higher average GDP compared to Africa, \~8x.
- Americas, Asia, and Europe are really closely clustered. They are all within less than \$2k of each other.
- I was surprised to see North America and South America as one continent. I think splitting them up would tell a more interesting story about the gap between the two hemispheres.

**Difficulties & Approaches**:

- I really wanted to be able to average the data in the plotting function, but it made it way too messy and I kept getting confused so I had to make a separate data frame
- Getting the graph to be visually appealing was a process. I first started by making the color better than the stock gray, then I went on to adding more informative axis titles, and I finished with adding a title. Then I also had to figure out how to center the title. Why not center it by default?? weird stuff

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
asia_min_year_sorted <- gapminder_min_year |> 
  filter(continent == "Asia") |> 
  arrange(desc(gdpPercap))

asia_max <- asia_min_year_sorted[1,1]

asia_max
asia_min_year_sorted



europe_min_year_sorted <- gapminder_min_year |> 
  filter(continent == "Europe") |> 
  arrange(desc(gdpPercap))

europe_max <- europe_min_year_sorted[1,1]

europe_max
europe_min_year_sorted



americas_min_year_sorted <- gapminder_min_year |> 
  filter(continent == "Americas") |> 
  arrange(desc(gdpPercap))

americas_max <- americas_min_year_sorted[1,1]

americas_max
americas_min_year_sorted

```

**Observations**:

- Identify the outlier countries from q2
  - The clearest outliers I saw were: Kuwait in Asia, Switzerland in Europe, and Canada + US in the Americas.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
countries_to_label <- c("Kuwait", "United States", "Switzerland", "Canada")

gapminder |>
  filter(year %in% c(year_min, year_max)) |>
  ggplot(
    aes(
      x = continent, 
      y = gdpPercap, 
      color = factor(year)
      )
    ) +
  geom_point() +
  geom_text(
    data = ~ filter(.x, country %in% countries_to_label),
    aes(label = country),
    vjust = -0.2,
    hjust = 1,
    show.legend = FALSE
  ) +
  scale_color_manual(
    name = "Year",
    values = c("1952" = "steelblue", "2007" = "firebrick")
  ) +
  labs(
    x = "Continent", 
    y = "Average GDP Per Capita ($USD)",
    title = "Average GDP Per Capita vs. Continent in 1952"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
  
```

**Observations**:

- Kuwait is the only country among the outliers that has a lower gdpPercap in 2007 than it did in 1952.
- Kuwait still has the highest gdpPercap in 2007 even though it fell by about half.
- Oceania used to have no outliers but now it does simply because there are so few countries.
- On average gdpPercap went up on all continents which tracks because of inflation and economic development.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
gapminder_min_year |> 
  ggplot() +
  geom_point(
    aes(
      x = gdpPercap,
      y = lifeExp
    )
  ) +
  labs(
    x = "Average GDP Per Capita ($USD)", 
    y = "Average Life Expectancy (Years)",
    title = "Average Life Expectancy vs. GDP Per Capita in 1952"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) 

gapminder_min_year |> 
  filter(country != "Kuwait") |> 
  ggplot() +
  geom_point(
    aes(
      x = gdpPercap,
      y = lifeExp
    )
  ) +
  labs(
    x = "Average GDP Per Capita ($USD)", 
    y = "Average Life Expectancy (Years)",
    title = "Average Life Expectancy vs. GDP Per Capita in 1952 (excluding Kuwait)"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) 
```

- Without eliminating the outlier of Kuwait, it is really difficult to see the relationships between gdpPercap and lifeExp.
- There is a wide spread but a clear positive trend between gdpPercap and lifeExp. Logically this makes sense given that a more resources lead to better healthcare, food, living conditions etc.
- While there is a positive trend, I wonder if there is a variable that has a stronger trend or a more clear trend with less spread.
- Also, what the heck happened in Kuwait for them to have such a high GDP?

```{r q5-task2}
coeff <- 1000  # rough ratio of GDP scale to life expectancy scale

gapminder |> 
  filter(country == "Kuwait") |>
  ggplot() +
  geom_line(
    aes(x = year, y = gdpPercap),
    color = "steelblue"
  ) + 
  geom_point(
    aes(x = year, y = gdpPercap),
    color = "steelblue"
  ) +
  geom_line(
    aes(x = year, y = lifeExp * coeff),
    color = "firebrick"
  ) + 
  geom_point(
    aes(x = year, y = lifeExp * coeff),
    color = "firebrick"
  ) +
  scale_y_continuous(
    name = "GDP per Capita",
    sec.axis = sec_axis(~ . / coeff, name = "Life Expectancy")
  )
```

- Despite the large variations and overall decline from 1950 to 2007 in Kuwait's GDP per capita, life expectancy is up from just over 50, to over 75 years.
- While on average a higher GDP is correlated with a higher life expectancy, it's not a fixed rule. Also, say with Kuwait for example, if the stock market crashes a lot of infrastructure still exists.

```{r q5-task3}

arranged_by_pop <- gapminder |> 
  filter(year == year_min) |> 
  arrange(pop, descending = TRUE)
arranged_by_pop


gapminder_min_year |> 
  ggplot() +
  geom_point(
    aes(
      x = continent,
      y = lifeExp
    )
  ) +
  labs(
    x = "Continent", 
    y = "Average Life Expectancy (Years)",
    title = "Average Life Expectancy vs. Continent in 1952"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) 

gapminder_min_year |> 
  filter(!country %in% c("India", "China", "United States")) |> 
  ggplot() +
  geom_point(
    aes(
      x = pop,
      y = lifeExp
    )
  ) +
  labs(
    x = "Population", 
    y = "Average Life Expectancy (Years)",
    title = "Average Life Expectancy vs. Population"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) 
```

- There seems to be no strong correlation between population and life expectancy.
- Again, the point about the outliers I made with Kuwait earlier stands. Outliers can't be discounted many times, but they make data much harder to process. So its good to look at data with and without an outlier.
