---
title: "Data Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 0
---

# Data: Basics

*Purpose*: When first studying a new dataset, there are very simple checks we
should perform first. These are those checks.

Additionally, we'll have our first look at the *pipe operator*, which will be
super useful for writing code that's readable.

*Reading*: (None)

```{r setup}
library(tidyverse)
```

```{r options, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

## First Checks
<!-- -------------------------------------------------- -->

### __q0__ Run the following chunk:

*Hint*: You can do this either by clicking the green arrow at the top-right of
the chunk, or by using the keybaord shortcut `Shift` + `Cmd/Ctrl` + `Enter`.

```{r }
head(iris)
```

This is a *dataset*; the fundamental object we'll study throughout this course.
Some nomenclature:

- The `1, 2, 3, ...` on the left enumerate the **rows** of the dataset
- The names `Sepal.Length`, `Sepal.Width`, `...` name the **columns** of the dataset
- The column `Sepal.Length` takes **numeric** values
- The column `Species` takes **string** values

### __q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?
Cut is how well the diamond has been cut/shaped. Color is the color of the diamond. Clarity is how clear the diamond is. 

*Hint*: You can run `?diamonds` to get information on a built-in dataset.

```{r load-packages, eval=FALSE}
library(tidyverse)
head(diamonds)
```

### __q2__ Run `glimpse(diamonds)`; what variables does `diamonds` have?
Vars:
- carat
- cut
- color
- clarity
- depth
- table
- price
- x
- y
- z
```{r}
glimpse(diamonds)
```


### __q3__ Run `summary(diamonds)`; what are the common values for each of the
variables? How widely do each of the variables vary?

*Hint*: The `Median` and `Mean` are common values, while `Min` and `Max` give us
a sense of variation.

```{r q3-task}
summary(diamonds)
```
Carat: Normal is 0.7-0.8, and Min is at 0.2 and max is at 5.01, however the IQR is only ~0.6.
Cut: Very good is the 'middle' value at 12k with the min at 1.6k and the max at 21.5k.
Color: Somewhat of an odd system. D is the best and J is the worst, however G, in the middle, is the highest number. The spread is from 11.3k to 2.8k which a normal value of around 8-10k. 
Clarity: Ranges from 13k to 2.5k with a normal value of around 5-9k. 
Depth: Mean and median are 61.75 and 61.8 respective;y and the IQR is 2.5 units which shows a narrow spread compared to the max and min spread of 36 (min at 43 and max at 79).
Table: Similar story to depth with a mean of 57.46 and an IQR of 3 with min at 43 and max at 95. Wide spread, but narrow IQR. 
Price: Wide range and wide IQR. Mean price was almost 4k with a median of 2.4k. The max and min spread was over 18k with an IQR of around 4.5k showing a varied dataset. 
X & Y: Both start at zero and have a mean of 5.73 with an IQR of about 1.5, however Y maxes at 58.9 while X tops out at just 10.74.
Z: Starts at zero, mean at 3.5 and max at 31.8 with an IQR of 1.1 meaning there is a tighter clump in this dimension compared to x and y. 


**Observations**:

- It is difficult to grade a diamonds quality by looking at the number associated with color given that the worst diamonds have a color of 2808 and the best at 6775 with a middle of the road diamond running 11292. No strictly increasing or decreasing relationship can be used for analysis here. 
- The X and Y dimensions of a diamond are incredibly similar for the average diamond (x mean= 5.731 y mean = 5.735) but are very different for outliers (x max = 10.74, y max = 58.9).
- High value diamonds skew the average mean which is shown by the median value of $2,401 and an average value of $3,933, over a 50% increase. 


You should always analyze your dataset in the simplest way possible, build
hypotheses, and devise more specific analyses to probe those hypotheses. The
`glimpse()` and `summary()` functions are two of the simplest tools we have.

## The Pipe Operator
<!-- -------------------------------------------------- -->

Throughout this class we're going to make heavy use of the *pipe operator*
`%>%`. This handy little function will help us make our code more readable.
Whenever you see `%>%`, you can translate that into the word "then". For
instance

```{r pipe-example}
diamonds %>%
  group_by(cut) %>%
  summarize(carat_mean = mean(carat))
```

Would translate into the tiny "story"

- Take the `diamonds` dataset, *then*
- Group it by the variable `cut`, *then*
- summarize it by computing the `mean` of `carat`

*What the pipe actually does*. The pipe operator `LHS %>% RHS` takes its
left-hand side (LHS) and inserts it as an the first argument to the function on
its right-hand side (RHS). So the pipe will let us take `glimpse(diamonds)` and
turn it into `diamonds %>% glimpse()`.

### __q4__ Use the pipe operator to re-write `summary(diamonds)`.

```{r q4-task}
diamonds %>% summary()
```

## Reading Data
<!-- -------------------------------------------------- -->

So far we've only been looking at built-in datasets. Ultimately, we'll want to read in our own data. We'll get to the art of loading and *wrangling* data later, but for now, know that the `readr` package provides us tools to read data. Let's quickly practice loading data below.

### __q5__ Use the function `read_csv()` to load the file `"./data/tiny.csv"`.

```{r q5-task}
read_csv("./data/tiny.csv")
df_q5
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data00-basics-assignment.Rmd).
